### 2024-06-17 sections/2-methods.tex
shrunk and combined algs and tables, still need to remove 50 lines of text.

Hereafter, when we refer to the LFD of a cluster, it is estimated at the length scale of the cluster radius and half that radius.
This allows us to rewrite Definition~\ref{eq:methods:lfd-original} as:

\begin{equation}
    \log_2 \left( \frac{|C(c, r)|}{|C(c, \frac{r}{2})|} \right)
    \label{eq:methods:lfd-simplified}
\end{equation}

where $|C(c, r)|$ is the cardinality of the cluster $C$ with center $c$ and radius $r$, and $C(c, \frac{r}{2})$ is the set of points in $C$ which are no more than a distance $\frac{r}{2}$ from $c$, i.e. $C(c, \frac{r}{2}) = \{p : p \in C \land f(c, p) \leq \frac{r}{2}\}$.

@@

To check whether $d \leq \rho$, we note that

\begin{equation*}
    d = \overline{m q' \vphantom{l}} = \overline{m r \vphantom{l}} - \overline{q' r \vphantom{l}} = \frac{\overline{l r}}{2} - \overline{q' r \vphantom{l}}.
\end{equation*}

Let $\theta$ denote $\angle l r q$, as shown in Figure~\ref{fig:methods:overlapping-children}.
By the Law of Cosines on $\triangle l r q$, we have that

\begin{equation*}
    \text{cos}(\theta) = \tfrac{\overline{l r}^2 + \ \overline{q r \vphantom{l}}^2 - \ \overline{q l}^2}{2 \cdot \overline{l r} \cdot \overline{q r \vphantom{l}}}.
\end{equation*}

Since $\triangle r q q'$ is a right triangle, we also have that

\begin{equation*}
    \text{cos}(\theta) = \tfrac{\overline{q' r \vphantom{l}}}{\overline{q r \vphantom{l}}}.
\end{equation*}

Combining the previous two equations and solving for $\overline{q' r \vphantom{l}}$, we have that

\begin{equation*}
    \overline{q' r \vphantom{l}} = \tfrac{\overline{q r \vphantom{l}}^2 + \ \overline{l r}^2 - \ \overline{q l}^2}{2 \cdot \overline{l r}}.
\end{equation*}

Substituting for $\overline{q' r \vphantom{l}}$ in the equation for $d$, we have that

\begin{equation*}
    d = \tfrac{\overline{l r}}{2} - \tfrac{\overline{q r \vphantom{l}}^2 + \ \overline{l r}^2 - \ \overline{q l}^2}{2 \cdot \overline{l r}} = \tfrac{\overline{q l}^2 - \overline{q r \vphantom{l}}^2}{2 \cdot \overline{l r}}.
\end{equation*}

Thus,

\begin{equation}
    d \leq \rho \iff (\overline{q l} + \overline{q r \vphantom{l}})(\overline{q l} - \overline{q r \vphantom{l}}) \leq 2 \cdot \overline{l r} \cdot \rho.
    \label{eq:methods:overlapping-children}
\end{equation}

Note, in particular, that Equation~\ref{eq:methods:overlapping-children} only requires distances between real points, and so it can be used with any distance function, even when $q'$ and $m$ are not real points or cannot be imputed from the data.

To perform $\rho$-NN search, we first perform a coarse \textit{tree-search}, as outlined in Algorithm~\ref{alg:methods:rnn-search:tree-search}, to find the leaf clusters which overlap with the query ball or any clusters which lie entirely within the query ball.
Then, for all such clusters, we perform a finer-grained \textit{leaf-search}, as outlined in Algorithm~\ref{alg:methods:rnn-search:leaf-search}, to find all points which are no more than a distance $\rho$ from the query.

\begin{algorithm} 
    \caption{tree-search($C$, $q$, $\rho$)} 
    \label{alg:methods:rnn-search:tree-search} 
    \begin{algorithmic}[0]
        \Require $C$, a cluster
        \Require $q$, a query
        \Require $\rho$, a search radius

@@

            \If{$(\overline{q l} + \overline{q r \vphantom{l}})(\overline{q l} - \overline{q r \vphantom{l}}) \leq 2 \cdot \overline{l r} \cdot \rho$}
                \State \textbf{return} tree-search($L, q, \rho$) $\cup$ tree-search($R, q, \rho$)
            \Else
                \State \textbf{return} tree-search($R, q, \rho$)
            \EndIf
        \EndIf
    \end{algorithmic}
\end{algorithm}

\begin{algorithm} 
    \caption{leaf-search($S$, $q$, $\rho$)} 
    \label{alg:methods:rnn-search:leaf-search} 
    \begin{algorithmic}[0]
        \Require $S$, a set of clusters
        \Require $q$, a query
        \Require $\rho$, a search radius

@@

        \State \textbf{return} $H$
    \end{algorithmic}
\end{algorithm}

\begin{algorithm} 
    \caption{$\rho$-NN-search($root$, $q$, $\rho$)} 
    \label{alg:methods:rnn-search} 
    \begin{algorithmic}
        \Require $root$, the root cluster
        \Require $q$, a query
        \Require $\rho$, a search radius

        \State $S$ $\Leftarrow$ tree-search($root$, $q$, $\rho$)
        \State $H$ $\Leftarrow$ leaf-search($S$, $q$, $\rho$)
        \State \textbf{return} $H$
    \end{algorithmic}
\end{algorithm}

The asymptotic complexity of $\rho$-NN is the same as in~\cite{ishaq2019clustered} and shown in Equation~\ref{eq:methods:rnn-search-complexity}.

@@

\begin{algorithm} % enter the algorithm environment
    \caption{Repeated $\rho$-NN($root$, $q$, $k$)} % give the algorithm a caption
    \label{alg:methods:repeated-rnn} % and a label for \ref{} commands later in the document
    \begin{algorithmic}[0] % enter the algorithmic environment
        \Require $root$, the root cluster
        \Require $q$, a query
        \Require $k$, the number of neighbors to find

@@

        \State $H \Leftarrow$ sort($\rho$-NN-search($root$, $q$, $r$))
        \State \textbf{return} $H[.. k]$
    \end{algorithmic}
\end{algorithm}

@@

The work in~\cite{yu2015entropy} showed that

\begin{equation*}
    \sum_{C \in S} |C| \leq \gamma  \left| B(q, \rho_k) \right| \left(\frac{\rho_k + 2 \cdot \hat{r}}{\rho_k} \right)^d
\end{equation*}

where $\gamma$ is a constant. 
By definition of $\rho_k$, we have that $|B(q, \rho_k)| = k$.
Thus,

\begin{equation*}
    \sum_{C \in S} |C| \leq \gamma k \left( 1 + 2 \cdot \frac{\hat{r}}{\rho_k} \right)^d
\end{equation*}

An estimate for $\rho_k$ is still needed. 
For this, we once again rely on the assumption that the query is drawn from the same distribution as the rest of the data, and thus the LFD at the query point is not significantly different from the LFD of nearby clusters.

We let $\hat{d}$ be the (harmonic) mean LFD of the clusters in $Q$.
While ordinarily we compute LFD by comparing cardinalities of two balls with two different radii centered at \textit{the same} point, in order to estimate $\rho_k$, we instead compare the cardinality of a ball \textit{around the query} of radius $\rho_k$ to the mean cardinality, $\hat{|C|}$, of clusters in $Q$ at a radius equal to the mean of their radii, $\hat{r}$.
We justify this approach by noting that, since the query is from the same distribution as the rest of the data, we could move from the query to the center of one of the nearby clusters without significantly changing our estimate of the LFD.

By Equation~\ref{eq:methods:lfd-original},

\begin{equation*}
    \hat{d} = \frac{\log{}\frac{\hat{|C|}}{k}}{\log{}\frac{\hat{r}}{\rho_k}}
\end{equation*}

We can rearrange this equation to get

\begin{equation*}
    \frac{\hat{r}}{\rho_k} = \left( \frac{\hat{|C|}}{k} \right)^{\hat{d}^{-1}}
\end{equation*}

Using this to simplify the term for leaf-search in Equation~\ref{eq:methods:rnn-search-complexity}, we get

\begin{equation*}
    k \left( 1 + 2 \cdot \left( \frac{\hat{|C|}}{k} \right) ^ {\hat{d}^{-1}} \right)^d.
\end{equation*}

@@

\begin{algorithm} % enter the algorithm environment
    \caption{Breadth-First Sieve($root$, $q$, $k$)} % give the algorithm a caption
    \label{alg:methods:bredth-first-sieve} % and a label for \ref{} commands later in the document
    \begin{algorithmic}[0] % enter the algorithmic environment
        \Require $root$, the root cluster
        \Require $q$, a query
        \Require $k$, the number of neighbors to find

@@

        \State \textbf{return} $Q$
    \end{algorithmic}
\end{algorithm}

@@

\begin{algorithm} 
    \caption{Depth-First Sieve($root$, $q$, $k$)} 
    \label{alg:methods:depth-first-sieve} 
    \begin{algorithmic}[0]
        \Require $root$, the root cluster
        \Require $q$, a query
        \Require $k$, the number of neighbors to find

        \While{$|H| < k$ \textbf{or} $H.peek.\delta \geq Q.peek.\delta^{-}$}
            \While{$\neg (Q.peek$ is a leaf)}
                \State{$C \Leftarrow Q.pop$, the closest cluster}
                \State{$[L, R] \Leftarrow$ children of $C$}
                \State{$Q.push(L)$}
                \State{$Q.push(R)$}
            \EndWhile
            \State{$leaf \Leftarrow Q.pop$}
            \For{$p \in leaf$}
                \State{$H.push(p)$}
            \EndFor
            \While{$|H| > k$}
                \State{$H.pop$}
            \EndWhile
        \EndWhile
        \State Return $H$
    \end{algorithmic}
\end{algorithm}

@@

We start with the assumption that the LFD near the query does not differ significantly from that of nearby clusters, and we consider leaf clusters with cardinalities near $k$.
Let $d$ be the LFD in this region.
Then the number of leaf-clusters in $Q$ is bounded above by $2d$, e.g. we could have a cluster overlapping the query ball at each end of each of $\lceil d \rceil$ mutually-orthogonal axes.
In the worst-case scenario for tree-search, these leaf clusters would all come from different branches of the tree, and so tree-search looks at $2 \cdot \lceil d \rceil \cdot \log \mathcal{N}_{\hat{r}}(X)$ clusters.
Thus, the asymptotic complexity is

\begin{equation*}
    \mathcal{T} \coloneqq \mathcal{O} \big( \lceil d \rceil \cdot \log \mathcal{N}_{\hat{r}}(X) \big).
\end{equation*}

For leaf-search, the output size and scaling factor are the same as in Repeated $\rho$-NN, and so the asymptotic complexity is

\begin{equation*}
    \mathcal{L} \coloneqq \mathcal{O} \left( k \cdot \bigg( 1 + 2 \cdot \Big( \frac{\hat{|C|}}{k} \Big) ^ {d^{-1}} \bigg)^d \right).
\end{equation*}

>>>>

Hereafter, when we refer to the LFD of a cluster, it is estimated at the length scale of the cluster radius and half that radius.
This allows us to rewrite Definition~\ref{eq:methods:lfd-original} as
$
    \log_2 \left( \frac{|C(c, r)|}{|C(c, \frac{r}{2})|} \right)
    \label{eq:methods:lfd-simplified}
$
where $|C(c, r)|$ is the cardinality of the cluster $C$ with center $c$ and radius $r$, and $C(c, \frac{r}{2})$ is the set of points in $C$ which are no more than a distance $\frac{r}{2}$ from $c$, i.e. $C(c, \frac{r}{2}) = \{p : p \in C \land f(c, p) \leq \frac{r}{2}\}$.

@@

To check whether $d \leq \rho$, we note that $d = \overline{m q' \vphantom{l}} = \overline{m r \vphantom{l}} - \overline{q' r \vphantom{l}} = \frac{\overline{l r}}{2} - \overline{q' r \vphantom{l}}$.

Let $\theta$ denote $\angle l r q$, as shown in Figure~\ref{fig:methods:overlapping-children}.
By the Law of Cosines on $\triangle l r q$, we have that $\text{cos}(\theta) = \tfrac{\overline{l r}^2 + \ \overline{q r \vphantom{l}}^2 - \ \overline{q l}^2}{2 \cdot \overline{l r} \cdot \overline{q r \vphantom{l}}}$.
Since $\triangle r q q'$ is a right triangle, we also have that $\text{cos}(\theta) = \tfrac{\overline{q' r \vphantom{l}}}{\overline{q r \vphantom{l}}}$.
Combining the previous two equations and solving for $\overline{q' r \vphantom{l}}$, we have that $\overline{q' r \vphantom{l}} = \tfrac{\overline{q r \vphantom{l}}^2 + \ \overline{l r}^2 - \ \overline{q l}^2}{2 \cdot \overline{l r}}$.
Substituting for $\overline{q' r \vphantom{l}}$ in the equation for $d$, we have that $d = \tfrac{\overline{l r}}{2} - \tfrac{\overline{q r \vphantom{l}}^2 + \ \overline{l r}^2 - \ \overline{q l}^2}{2 \cdot \overline{l r}} = \tfrac{\overline{q l}^2 - \overline{q r \vphantom{l}}^2}{2 \cdot \overline{l r}}$.

Thus, $d \leq \rho \iff (\overline{q l} + \overline{q r \vphantom{l}})(\overline{q l} - \overline{q r \vphantom{l}}) \leq 2 \cdot \overline{l r} \cdot \rho$. Note, in particular, that this only requires distances between real points, and so it can be used with any distance function, even when $q'$ and $m$ are not real points or cannot be imputed from the data.

To perform $\rho$-NN search, we first perform a coarse \textit{tree-search}, as outlined in Algorithm~\ref{alg:methods:rnn-search:tree-search}, to find the leaf clusters which overlap with the query ball or any clusters which lie entirely within the query ball.
Then, for all such clusters, we perform a finer-grained \textit{leaf-search}, as outlined in Algorithm~\ref{alg:methods:rnn-search:leaf-search}, to find all points which are no more than a distance $\rho$ from the query.
The asymptotic complexity of $\rho$-NN is the same as in~\cite{ishaq2019clustered} and shown in Equation~\ref{eq:methods:rnn-search-complexity}.

\begin{minipage}{0.6\textwidth}
\begin{algorithm}[H]
    \caption{tree-search($C$, $q$, $\rho$)}
    \label{alg:methods:rnn-search:tree-search}
    \begin{tiny}
    \begin{algorithmic}[0]
        \Require $C$, a cluster
        \Require $q$, a query
        \Require $\rho$, a search radius

@@

            \If{$(\overline{q l} + \overline{q r \vphantom{l}})(\overline{q l} - \overline{q r \vphantom{l}}) \leq 2 \cdot \overline{l r} \cdot \rho$}
                \State \textbf{return} tree-search($L, q, \rho$) $\cup$ tree-search($R, q, \rho$)
            \Else
                \State \textbf{return} tree-search($R, q, \rho$)
            \EndIf
        \EndIf
    \end{algorithmic}
    \end{tiny}
\end{algorithm}
\end{minipage}
\hfill
\begin{minipage}{0.35\textwidth}
\begin{algorithm}[H]
    \caption{leaf-search($S$, $q$, $\rho$)}
    \label{alg:methods:rnn-search:leaf-search}
    \begin{tiny}
    \begin{algorithmic}[0]
        \Require $S$, a set of clusters
        \Require $q$, a query
        \Require $\rho$, a search radius

@@

        \State \textbf{return} $H$
    \end{algorithmic}
    \end{tiny}
\end{algorithm}
\end{minipage}

\begin{algorithm}
    \caption{$\rho$-NN-search($root$, $q$, $\rho$)}
    \label{alg:methods:rnn-search}
    \begin{tiny}
    \begin{algorithmic}
        \Require $root$, the root cluster
        \Require $q$, a query
        \Require $\rho$, a search radius

        \State $S$ $\Leftarrow$ tree-search($root$, $q$, $\rho$)
        \State $H$ $\Leftarrow$ leaf-search($S$, $q$, $\rho$)
        \State \textbf{return} $H$
    \end{algorithmic}
    \end{tiny}
\end{algorithm}

@@

\begin{algorithm} % enter the algorithm environment
    \caption{Repeated $\rho$-NN($root$, $q$, $k$)} % give the algorithm a caption
    \label{alg:methods:repeated-rnn} % and a label for \ref{} commands later in the document
    \begin{tiny}
    \begin{algorithmic}[0] % enter the algorithmic environment
        \Require $root$, the root cluster
        \Require $q$, a query
        \Require $k$, the number of neighbors to find

@@

        \State $H \Leftarrow$ sort($\rho$-NN-search($root$, $q$, $r$))
        \State \textbf{return} $H[.. k]$
    \end{algorithmic}
    \end{tiny}
\end{algorithm}

@@

The work in~\cite{yu2015entropy} showed that $\sum_{C \in S} |C| \leq \gamma  \left| B(q, \rho_k) \right| \left(\frac{\rho_k + 2 \cdot \hat{r}}{\rho_k} \right)^d$, where $\gamma$ is a constant.
By definition of $\rho_k$, we have that $|B(q, \rho_k)| = k$.
Thus, $\sum_{C \in S} |C| \leq \gamma k \left( 1 + 2 \cdot \frac{\hat{r}}{\rho_k} \right)^d$.
However, an estimate for $\rho_k$ is still needed.
For this, we once again rely on the assumption that the query is drawn from the same distribution as the rest of the data, and thus the LFD at the query point is not significantly different from the LFD of nearby clusters.
We let $\hat{d}$ be the (harmonic) mean LFD of the clusters in $Q$.
While ordinarily we compute LFD by comparing cardinalities of two balls with two different radii centered at \textit{the same} point, in order to estimate $\rho_k$, we instead compare the cardinality of a ball \textit{around the query} of radius $\rho_k$ to the mean cardinality, $\hat{|C|}$, of clusters in $Q$ at a radius equal to the mean of their radii, $\hat{r}$.
We justify this approach by noting that, since the query is from the same distribution as the rest of the data, we could move from the query to the center of one of the nearby clusters without significantly changing our estimate of the LFD.
By Equation~\ref{eq:methods:lfd-original}, $\hat{d} = \frac{\log{}\frac{\hat{|C|}}{k}}{\log{}\frac{\hat{r}}{\rho_k}}$.
We can rearrange this equation to get $\frac{\hat{r}}{\rho_k} = \left( \frac{\hat{|C|}}{k} \right)^{\hat{d}^{-1}}$.
Using this to simplify the term for leaf-search in Equation~\ref{eq:methods:rnn-search-complexity}, we get $k \left( 1 + 2 \cdot \left( \frac{\hat{|C|}}{k} \right) ^ {\hat{d}^{-1}} \right)^d$.

@@

\begin{minipage}{0.46\textwidth}
\begin{algorithm}[H] % enter the algorithm environment
    \caption{Breadth-First Sieve($root$, $q$, $k$)} % give the algorithm a caption
    \label{alg:methods:bredth-first-sieve} % and a label for \ref{} commands later in the document
    \begin{tiny}
    \begin{algorithmic}[0] % enter the algorithmic environment
        \Require $root$, the root cluster
        \Require $q$, a query
        \Require $k$, the number of neighbors to find

@@

        \State \textbf{return} $Q$
    \end{algorithmic}
    \end{tiny}
\end{algorithm}
\end{minipage}
\hfill
\begin{minipage}{0.46\textwidth}
\begin{algorithm}[H]
    \caption{Depth-First Sieve($root$, $q$, $k$)}
    \label{alg:methods:depth-first-sieve}
    \begin{tiny}
    \begin{algorithmic}[0]
        \Require $root$, the root cluster
        \Require $q$, a query
        \Require $k$, the number of neighbors to find

        \While{$|H| < k$ \textbf{or} $H.peek.\delta \geq Q.peek.\delta^{-}$}
            \While{$\neg (Q.peek$ is a leaf)}
                \State{$C \Leftarrow Q.pop$, the closest cluster}
                \State{$[L, R] \Leftarrow$ children of $C$}
                \State{$Q.push(L)$}
                \State{$Q.push(R)$}
            \EndWhile
            \State{$leaf \Leftarrow Q.pop$}
            \For{$p \in leaf$}
                \State{$H.push(p)$}
            \EndFor
            \While{$|H| > k$}
                \State{$H.pop$}
            \EndWhile
        \EndWhile
        \State Return $H$
    \end{algorithmic}
    \end{tiny}
\end{algorithm}
\end{minipage}

@@

@@

We start with the assumption that the LFD near the query does not differ significantly from that of nearby clusters, and we consider leaf clusters with cardinalities near $k$.
Let $d$ be the LFD in this region.
Then the number of leaf-clusters in $Q$ is bounded above by $2d$, e.g. we could have a cluster overlapping the query ball at each end of each of $\lceil d \rceil$ mutually-orthogonal axes.
In the worst-case scenario for tree-search, these leaf clusters would all come from different branches of the tree, and so tree-search looks at $2 \cdot \lceil d \rceil \cdot \log \mathcal{N}_{\hat{r}}(X)$ clusters.
Thus, the asymptotic complexity is $\mathcal{T} \coloneqq \mathcal{O} \big( \lceil d \rceil \cdot \log \mathcal{N}_{\hat{r}}(X) \big)$.
For leaf-search, the output size and scaling factor are the same as in Repeated $\rho$-NN, and so the asymptotic complexity is $\mathcal{L} \coloneqq \mathcal{O} \left( k \cdot \bigg( 1 + 2 \cdot \Big( \frac{\hat{|C|}}{k} \Big) ^ {d^{-1}} \bigg)^d \right)$.
